---
title: "Introduction to ESCO"
author: "Jinjin Tian"
date: "`r Sys.Date()`"
output:
    BiocStyle::html_document:
        toc: true
vignette: >
  %\VignetteIndexEntry{An introduction to the ESCO simulator}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r knitr-options, echo = FALSE, message = FALSE, warning = FALSE}
# To render an HTML version that works nicely with github and web pages, do:
# rmarkdown::render("vignettes/escoter.Rmd", "all")
knitr::opts_chunk$set(fig.align = 'center', 
                      fig.width = 10, fig.height = 8,
                      dev = 'png')

# Use exact BSPARAM to avoid warnings
options(BiocSingularParam.default = BiocSingular::ExactParam())
setwd("~/SplatterESCO")
devtools::load_all()
showdir = "./show"
if(!dir.exists(showdir))dir.create(showdir)
```

Welcome to ``SplatterESCO``! ``SplatterESCO`` is an R package for the simple simulation of
single-cell RNA sequencing data, built upon the `splatter` package, with newly added methods ESCO for simulating gene co-expression. This vignette gives an overview and introduction to ESCO's functionality.

# Installation

``SplatterESCO`` can be download from Github:
         https://github.com/JINJINT/SplatterESCO

# Overview

We use the term 'ESCO' to refer
to the newly proposed Ensemble single cell simulator with gene co-expression, and differentiate it from the package SplatterESCO itself. As for the marginal behaviour of genes and cells, the core of the ESCO model is foundimentally the same with the original Splat: they both use a gamma-Poisson distribution with a enforced mean-variance trend to generate gene counts, and use a log-normal/normal distribution to give an expected library size for each cell, while outlier genes (genes with mean expression outside the gamma
distribution) and dropouts (random knock out of counts based on mean expression) are introduced along the way. 

ESCO can also simulate differential expression between groups of different
types of cells or differentiation paths between different cells types where
expression changes in a continuous way. These are described further in the [ESCO simulation] section.


# ESCO simulation
Once we have a set of parameters we are happy with, we can use `escoSimulate`
to simulate counts. If we want to make small adjustments to the parameters we
can provide them as additional arguments, alternatively if we don't supply any
parameters the defaults will be used:

```{r escoSimulate, warning = FALSE}
sim <- escoSimulate(nGenes = 100, nCells  = 50)
sim
# Access the counts
assays(sim)$TrueCounts[1:5, 1:5]
# Information about genes
head(rowData(sim))
# Information about cells
head(colData(sim))
# Information about paramters
str(metadata(sim)$Params)
```

Looking at the output of `escoSimulate` we can see that `sim` is
`SingleCellExperiment` object contains phenotype information about
each cell (accessed using `colData`); feature information about each gene (accessed using `rowData`); genes by cells data matrix and intermediate values of the simulation (accessed using `assays`); and parameters used to generate this simulation (accessed using `metadata`).

The main part of this object is a series of features
by samples matrix containing the simulated true counts (accessed using `assays(sim)$TrueCounts`); counts with zero-inflation noise; (accessed using `assays(sim)$counts`); counts with downsample noise ((accessed using `assays(sim)$observedcounts`). By default, all three kinds of data matrix will be simulated, while users are allowed to simulated only one or two of them for time/space saving, via specifying the ``dropout.type`` in parameters setting.

For more details about the `SingleCellExperiment` object refer to the [vignette][SCE-vignette]. For information about what you can do with `scater` refer to the `scater` documentation and [vignette][scater-vignette].

Additionally, in order to have better reproducibility, ESCO also allows multiple trials of noisy data generated from one common true counts in one single call, and save them seperately in the directory user provided. The returned ``sim`` on the other hand will only contain one trial of noisy data in order to avoid memory exhuastion.

```{r escoSimulate1, warning = FALSE}
if(!dir.exists(paste0(showdir,"case1")))
  dir.create(paste0(showdir,"case1"))
sim <- escoSimulate(nGenes = 100, nCells  = 50, 
                    trials = 5, 
                    dirname = paste0(showdir,"case1/"), 
                    verbose = FALSE)
list.files(path = paste0(showdir,"case1/"), full.names = FALSE)
```

Moreover, ESCO also allows noisy data of different configuration of noise level data generated from one common true counts in one single call, and save them seperately in the directory user provided. The returned ``sim`` on the other hand will only contain one configuration of noisy data in order to avoid memory exhuastion.

```{r escoSimulate2, warning = FALSE}
if(!dir.exists(paste0(showdir,"case2")))dir.create(paste0(showdir,"case2"))
sim <- escoSimulate(nGenes = 100, nCells  = 50, 
                    dropout.mid = c(1,2), 
                    alpha_mean = c(0.1, 0.2), 
                    trials = 2, 
                    dirname = paste0(showdir,"case2/"), 
                    verbose = FALSE)
list.files(path = paste0(showdir,"case2/"), full.names = FALSE)
```


## Simulating one cell group
In this section, we will show specific examples of how to simulate one homogenrous cell with/ without gene co-expression group using ESCO. Fisrtly, we show exmaple of simulation without gene co-expression, where ``heatdata`` and ``heatgcn`` are newly added functions for convinient visualization. 

```{r simulate single random corr, warning = FALSE, fig.height = 4, fig.width = 10}
sim <- escoSimulateSingle(nGenes = 100, nCells = 50, 
                          lib.loc = 7, withcorr = TRUE, verbose = FALSE)

# get the data
datalist = list("simulated truth" = assays(sim)$TrueCounts, 
                "zero-inflated" = assays(sim)$counts, 
                "down-sampled" = assays(sim)$observedcounts)

# plotting the data
heatdata(datalist, norm = FALSE, size = 2, ncol = 3)
```

```{r, warning = FALSE, fig.height = 4, fig.width = 10}
# plotting the GCN
rholist = metadata(sim)$Params@corr
corrgenes = rownames(rholist[[1]])
gcnlist = lapply(datalist, function(data)gcn(data, genes = corrgenes))
gcnlist = append(gcnlist, list("given truth" = rholist[[1]]), 1)
heatgcn(gcnlist, size = 3, ncol = 4)
```

You can also provide your own defined GCN. In the following, we take the randomly generated correlation structure in the last simulation as our input.

```{r simulate single given corr, warning = FALSE, fig.height = 4, fig.width = 10}
sim <- escoSimulateSingle(nGenes = 100, nCells = 50, 
                          lib.loc = 7, withcorr = TRUE, 
                          corr = rholist, verbose = FALSE)

# plotting the data
datalist = list("simulated truth" = assays(sim)$TrueCounts, 
              "zero-inflated" = assays(sim)$counts, 
              "down-sampled" = assays(sim)$observedcounts)
heatdata(datalist, norm = FALSE, size = 2, ncol = 3)
```

```{r, warning  = FALSE, fig.height = 3, fig.width = 10}
# plotting the GCN
rholist = metadata(sim)$Params@corr
corrgenes = rownames(rholist[[1]])
gcnlist = lapply(datalist, function(data)gcn(data, genes = corrgenes))
gcnlist = append(gcnlist, list("given truth" = rholist[[1]]), 1)
heatgcn(gcnlist, size = 2, ncol = 4)
```


## Simulating multiple discrete groups

So far we have only simulated a single population of cells but often we are
interested in investigating a mixed population of cells and looking to see what
cell types are present or what differences there are between them. ESCO is
able to simulate these situations by changing the `method` argument. Here we are
going to simulate two groups, by specifying the `group.prob`, ``de.prob`` parameters. The gene-gene correlation can be generated randomly (by default), or given by users.

```{r groups, warning = FALSE, fig.height = 6, fig.width = 10}
sim<-escoSimulateGroups(nGenes = 200, nCells = 100,
                        group.prob = c(0.6, 0.4), deall.prob = 0.3,
                        de.prob = c(0.3, 0.7),
                        de.facLoc = c(1.9, 2.5), withcorr = TRUE, 
                        trials = 1, verbose =FALSE)

# organize the marker gene info
genegroup = paste0("Group", rowData(sim)$GeneGroup)
genegroup[which(genegroup=="Group0")] = "None"
geneinfo = data.frame(genes = rowData(sim)$Gene, 
                      newcelltype = as.factor(genegroup))

# organize the cell info
cellinfo = data.frame(çells = colData(sim)$Cell, 
                       newcelltype= as.factor(colData(sim)$Group))

# get the data
datalist = list("simulated truth" = assays(sim)$TrueCounts, 
                "zero-inflated" = assays(sim)$counts, 
                "down-sampled" = assays(sim)$observedcounts)

# plotting the data                     
heatdata(datalist, cellinfo = cellinfo, geneinfo = geneinfo, 
         size = 1, ncol = 3)
```

```{r, warning = FALSE, fig.height = 4, fig.width = 10}
# plot GCN for all marker genes (i.e. DE genes) across all cell groups
degeneinfo = geneinfo[which(geneinfo$newcelltype!="None"),]
degeneinfo$newcelltype = droplevels(degeneinfo$newcelltype)
degcnlist = lapply(datalist, function(data)gcn(data, genes = degeneinfo$genes))
heatgcn(degcnlist, geneinfo = degeneinfo, size = 2, ncol = 3)
```

```{r, warning = FALSE, fig.height = 4, fig.width = 10}
# plot GCN for marker gene within one cell group
rholist = metadata(sim)$Params@corr
group2_gcnlist = lapply(datalist, 
                        function(data){
                          gcn(data[,which(colData(sim)$Group=="Group2")], 
                              CPM2 = TRUE, 
                              genes = rownames(rholist[["Group2"]]))})
group2_gcnlist = append(group2_gcnlist, 
                        list("given truth" = rholist[["Group2"]]), 1)
heatgcn(group2_gcnlist, size = 3, ncol = 4)
```


## Simulating tree structured cell group

```{r tree, warning  = FALSE}
#generate the tree
yaml="
    name: All
    Group1:
        Group1-1:
             params: NULL
    Group2:
        Group2-1:
             params: NULL
        Group2-2:
             params: NULL
"
os.list = yaml::yaml.load(yaml)
tree = data.tree::as.Node(os.list)
tree = data.tree::as.phylo.Node(tree)
```

```{r, warning  = FALSE, fig.height = 7, fig.width = 10}
# simulation
sim<-escoSimulateTree(nGenes = 500, nCells = 300, 
                      tree = list(tree), 
                      group.prob = c(0.4, 0.3, 0.3), 
                      deall.prob = 0.3, 
                      de.center = 2,
                      de.prob = c(0.5, 0.5, 0.5),
                      de.facLoc = c(1.5, 2, 3), 
                      de.facScale = c(0.9, 0.9, 0.9), 
                      withcorr = TRUE,  
                      trials = 1, verbose = FALSE)
# get the data
datatrue = assays(sim)$TrueCounts

# get the cellinfo
cellinfo  = data.frame(cell = colData(sim)$Cell, 
                       newcelltype = as.factor(colData(sim)$Group))
levels(cellinfo$newcelltype) = tree$tip.label

# get the geneinfo
genegroup = paste0("Group", rowData(sim)$GeneGroup)
genegroup[which(genegroup=="Group0")] = "None"
geneinfo = data.frame(genes = rowData(sim)$Gene, 
                      newcelltype = as.factor(genegroup))
levels(geneinfo$newcelltype)[1:3] = tree$tip.label

# get the DE geneinfo
groups <- colData(sim)$Group
group.names <- sort(unique(groups))
group.facs.gene <- rowData(sim)[, paste0("DEFac", group.names)]
DEgene.name = as.character(rowData(sim)$Gene[which(group.facs.gene[,1]>1)])
degeneinfo = geneinfo[match(DEgene.name, geneinfo$genes),]

# plot the data
heatdata(list(datatrue), 
         colv = TRUE, 
         cellinfo = cellinfo, 
         geneinfo = degeneinfo, 
         genes = degeneinfo$genes,
         size = 1.5, ncol = 1)

```

## Simulating continous cell trajectory

The other situation that is often of interest is a differentiation process where
one cell type is changing into another. escoter approximates this process by
simulating a series of steps between two groups and randomly assigning each
cell to a step. We can create this kind of simulation using the `"paths"`
method.

```{r paths, warning=FALSE, fig.height = 3, fig.width = 4}
sim <- escoSimulatePath(nGenes = 1000, nCells = 500,
                        paths.deprob = 0.3,
                        paths.design = data.frame(
                          Path = c(1, 2, 3),
                          From = c(0, 1, 2),
                          Steps = c(100, 100, 100)
                        ),
                        cells.design = data.frame(
                          Path = c(1, 2, 3),
                          Probability = c(0.3, 0.3, 0.4),
                          Alpha = 1,
                          Beta = 1
                        ),
                        withcorr = TRUE,verbose = FALSE)


datatrue = assays(sim)$TrueCounts

# get the cellinfo
cellinfo = data.frame(cell = colData(sim)$Cell, 
                      newcelltype = colData(sim)$Path)
# get the pesudo time
celltime = data.frame(path = as.numeric(colData(sim)$Path), 
                      step = as.numeric(colData(sim)$Step))
celltime = order(celltime[,1], celltime[,2])

# get the geneinfo
degenes = which(metadata(sim)$Params@paths.DEgenes==1)

# plot the data
umapplot(t(t(datatrue)/colSums(datatrue)), 
         celltype  = colData(sim)$Path, 
         labels = levels(as.factor(colData(sim)$Path)))
```

```{r, warning  = FALSE}
heatdata(list("simulated truth" = datatrue[degenes,]),
         cellinfo = cellinfo,
         colv = celltime, size = 1, ncol = 1)
```


# GCN comparison
The new package also provide simple function to compare the GCN of a list of  data, using the first one in the list as the benchmark. You can choose different correlation metric; error metric; clustering method etc. 

```{r, warning  = FALSE, eval = FALSE}
gcnerr = evaluation(datalist, genes, cells, 
                    corrname = c("pearson", "spearman"), 
                    CPM2  = CPM2, errname = c("MSE", "ARI", "AUC"), 
                    clustk = c(2:9), clustmethod = "hclust")

```




[gamma]: https://en.wikipedia.org/wiki/Gamma_distribution
[poisson]: https://en.wikipedia.org/wiki/Poisson_distribution
[scater-vignette]: https://bioconductor.org/packages/release/bioc/vignettes/scater/inst/doc/vignette.html
[SCE-vignette]: https://bioconductor.org/packages/devel/bioc/vignettes/SingleCellExperiment/inst/doc/intro.html
